---
title: "Fundamentals"
---

## Overview

This section covers some of the most fundamental operations of both languages. These include <span style="color:#077DC2">variable</span>/<span style="color:#FF9B00">object</span> assignment, data <span style="color:#077DC2">type</span>/<span style="color:#FF9B00">class</span>, arithmetic, etc. External data are _not_ included in this page.

Note that any line in a code chunk preceded by a hashtag (`#`) is a "comment" and is not evaluated in either language. Including comments is generally good practice because it allows humans to read and understand code that may otherwise be unclear to them.

## Library Loading

This section assumes you've already installed the needed python / R packages. See the chunks in the home page of this website if you need a refresher on which packages are needed. It is not enough to install a package onto a given computer, you also have to specifically load that library into your working environment for its functions to be available to you in a specific script / document.

:::panel-tabset
## `r fontawesome::fa(name = "r-project", fill = "#FF9B00", a11y = "sem")` R

R libraries are loaded with the `library` function

```{r r-libs, warning = F, message = F}
library(tidyverse)
```

## `r fontawesome::fa(name = "python", fill = "#077DC2", a11y = "sem")` Python

python libraries are loaded with the `import` statement and can be aliased into simpler names with `as` at the same time.

```{python py-libs}
import numpy as np
```
:::

## Assignment

At its most basic, we want to store data in code in such a way that we can use / manipulate it via our scripts. This requires **assigning** data to a <span style="color:#077DC2">variable</span>/<span style="color:#FF9B00">object</span> with the **assignment operator**.

:::panel-tabset
## `r fontawesome::fa(name = "r-project", fill = "#FF9B00", a11y = "sem")` R

In R, the assignment operator is `<-`. To use it, the name of the new object-to-be is on the left of the arrow and the information to assign is on the right.

```{r r-assign}
# Make a simple object
a <- 2

# Check it out
a
```

## `r fontawesome::fa(name = "python", fill = "#077DC2", a11y = "sem")` Python

In Python, the assignment operator is `=`. To use it, the name of the new object-to-be is on the left of the equal sign and the information to assign is on the right.

```{python py-assign}
# Make a simple object
a = 2

# Check it out
a
```
:::

Once we've created a <span style="color:#077DC2">variable</span>/<span style="color:#FF9B00">object</span> we can then use the information stored inside of it in downstream operations! For example, we could perform basic arithmetic on our <span style="color:#077DC2">variable</span>/<span style="color:#FF9B00">object</span> and assign the result to a new <span style="color:#077DC2">variable</span>/<span style="color:#FF9B00">object</span>.

:::panel-tabset
## `r fontawesome::fa(name = "r-project", fill = "#FF9B00", a11y = "sem")` R

Addition, subtraction, multiplication, and division share operators across both languages (`+`, `-`, `*`, and `/` respectively). However, in R exponents use `^`.

```{r r-math}
# Raise to an exponent
b <- a^3

# Check out the result
b
```

## `r fontawesome::fa(name = "python", fill = "#077DC2", a11y = "sem")` Python

Addition, subtraction, multiplication, and division share operators across both languages (`+`, `-`, `*`, and `/` respectively). However, in Python exponents use `**`

```{python py-math}
# Raise to an exponent
b = a**3

# Check out the result
b
```
:::

## <span style="color:#077DC2">Type</span> & <span style="color:#FF9B00">Class</span>

Some operations are only possible on some categories of information. For instance, we can only perform arithmetic on numbers. In Python this is known as the <span style="color:#077DC2">variable</span>'s <span style="color:#077DC2">type</span> & while in R this is the <span style="color:#FF9B00">object</span>'s <span style="color:#FF9B00">class</span>. In either case, it's important to know--and be able to check--this information about the <span style="color:#077DC2">variables</span>/<span style="color:#FF9B00">objects</span> with which we are working.

:::panel-tabset
## `r fontawesome::fa(name = "r-project", fill = "#FF9B00", a11y = "sem")` R

In R we use the `class` function to get this information. Note that the names of R classes sometimes differ from their equivalents in Python.

```{r r-class1}
# Check class of an integer
class(37)
```

```{r r-class2}
# Check class of a decimal
class(3.14159)
```

```{r r-class3}
# Check class of text
class("my hands are typing words")
```

## `r fontawesome::fa(name = "python", fill = "#077DC2", a11y = "sem")` Python

In python, the `type` function returns the type of the data object. Note that the names of Python types sometimes differ from their equivalents in R.

```{python py-class1}
# Check type of an integer
type(37)
```

```{python py-class2}
# Check type of a decimal
type(3.14159)
```

```{python py-class3}
# Check type of text
type("my hands are typing words")
```
:::

## Indexing

When our <span style="color:#077DC2">variables</span>/<span style="color:#FF9B00">objects</span> have more than one <span style="color:#077DC2">item</span>/<span style="color:#FF9B00">element</span> we may want to examine the piece of information at a specific position. This position is the "index position" and can be accessed in either language fairly easily.

In order to explore this more fully, let's make some example multi-component <span style="color:#077DC2">variables</span>/<span style="color:#FF9B00">objects</span>.

:::panel-tabset
## `r fontawesome::fa(name = "r-project", fill = "#FF9B00", a11y = "sem")` R

In R, one of the fundamental data structures is a "vector". Vectors are assembled with the concatenation function (`c`) where each item is separated by commas (`,`) and the set of them is wrapped in parentheses (`(...)`).

Note that the class of the object comes from the vector's _contents_ rather than the fact that it is a vector. All elements in a vector therefore must share a class.

```{r r-index-prep}
# Make a multi-item variable
x <- c(1, 2, 3, 4, 5)

# Check it out
class(x)
```

## `r fontawesome::fa(name = "python", fill = "#077DC2", a11y = "sem")` Python

In Python the fundamental data structure is a "list". Lists are assembled either by wrapping the items to include in square brackets (`[...]`) or by using the `list` function. In either case, each item is separated from the others by commas (`,`).

Note that the type of the variable comes from the _list itself_ rather than its contents. Lists therefore support items of multiple different types.

```{python py-index-prep}
# Make a multi-item variable
x = [1, 2, 3, 4, 5]

# Check it out
type(x)
```
:::

One crucial difference between R and Python is that Python is "0-based" meaning that the first <span style="color:#077DC2">item</span> is at index position `0` while in R the position of the equivalent <span style="color:#FF9B00">element</span> is `1`.

Fortunately, in either language the syntax for indexing is the same.

:::panel-tabset
## `r fontawesome::fa(name = "r-project", fill = "#FF9B00", a11y = "sem")` R

To index a multi-element object, simply append square brackets to the end of the object name and specify the number of the index position in which you are interested.

```{r r-index}
# Access the first element of the vector
x[1]
```

## `r fontawesome::fa(name = "python", fill = "#077DC2", a11y = "sem")` Python

To index a multi-item variable, simply append square brackets to the end of the variable name and specify the number of the index position in which you are interested.

```{python py-index}
# Access the first item of the list
x[0]
```
:::

